Lower Bounds on the Minimax Risk for the \\ Source Localization Problem

### 2017-01-24 paper.tex
Complete incorporating all edits.

@@

The sensors sample the continuous-space shifted impulse response, with some
additive noise. We denote the noiseless sampled version of $x(s; \theta)$ by
the $m$-length vector $\v x(\theta)$:
\begin{equation} \label{eq:sampled-signal}
	\v x(\theta) = \bigg[x(0; \theta), \ldots, x\Big(\frac{kS}{m}; \theta\Big), \ldots, x\Big(\frac{(m-1)S}{m}; \theta\Big)\bigg]^T
\end{equation}
where $m$ is the number of sensors and $k \in \{0, 1, \ldots, m-1\}$. The
additive noise is given by $\v \epsilon$ (to be described shortly), and the
noisy samples are represented by $\v y$:
\begin{equation} \label{eq:sensor-obs}
	\v y = \v x(\theta) + \v \epsilon
\end{equation}
In the source localization problem, we receive $n$ trials\footnote{The word
``trials'' comes from neuroscience literature, where several trials of an
experiment are run with a single participant. The participant is given the same
stimulus in each trial, evoking the same response in the brain (usually with
some trial-to-trial variability). For our purposes, we assume that trials are
effectively i.i.d. realizations of sensor noise, with no trial-to-trial
variability.} (i.e., $n$ realizations) of $\v y$ according to
equation~\eqref{eq:sensor-obs}, for the same point source $\theta$, using which
we need to estimate the location of the source.

@@

The noise $\v \epsilon$, introduced in equation~\eqref{eq:sensor-obs}, is the
only source of uncertainty in the problem. For our purposes, we assume that $\v
\epsilon$ is zero-mean and Gaussian, $\v \epsilon \sim \mathcal{N}(\v 0, \m
\Sigma)$. We restrict our analysis to the ``sensor noise'' setting, wherein
$\m\Sigma = \sigma^2 \m I$, i.e. the sensors are afflicted by i.i.d.\ noise.

@@

We start by lower bounding the minimax risk of the estimation problem with the
risk incurred in a multiple hypothesis testing problem. For this, we first need
to define a $2\delta$-packing:%
\begin{definition}
	A set $\Theta_{\mathcal{V}} = \{ \theta_v : v \in \mathcal{V} \}$ for some
	finite index set $\mathcal{V} \subset \mathbb N$ is said to be a
	$2\delta$-packing in the $\rho$-metric if
	\begin{equation}
		\rho(\theta_i, \theta_j) \geq 2\delta \qquad \forall \;\; \theta_i, \theta_j \in \Theta_{\mathcal{V}}.
	\end{equation}
\end{definition}
\begin{theorem} \label{thm:est-to-testing}%
	If we can find a $2\delta$-packing $\Theta_{\mathcal{V}}$ of $\Theta$, then
	we can lower bound the minimax estimation risk by the average testing risk:
	\begin{equation}
		\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) \inf_\psi \mathbb P (\psi(Y^n) \neq V)
	\end{equation}
	where $V$ is the unknown true hypothesis which takes values uniformly from
	$\mathcal{V}$, and $\psi$ is our estimate of the hypothesis.
\end{theorem}
For a proof of this theorem, we refer the reader to Proposition~13.3
in~\cite{Duchi2015Information}.  Intuitively, Theorem~\ref{thm:est-to-testing}
says that the error in estimating $\theta_i$ is likely to be more if it is
difficult to distinguish $\theta_i$ from $\theta_j$, i.e.\ if the probability
of error is high.

@@

Recall that we are interested in computing lower bounds on the minimax risk for
the source localization problem defined in
Section~\ref{sec:source-localization}. We measure \emph{risk} in terms of the
squared-distance between the true location and the estimated location. Hence,
our loss function is $\Phi(\rho(\theta, \hat\theta)) = \abs{\theta -
\hat\theta}^2$, where $\hat\theta$ is some estimate of the location based on
$n$ trials of the noisy sensor observations $\v y$.

We now state the main result of this paper.
\begin{theorem} \label{thm:main-lb}
	For a source localization problem as defined in
	section~\ref{sec:source-localization} with sensor noise and a spatial
	impulse response that is $\kappa$-Lipschitz continuous and has restricted
	support of width $w$, the \textcolor{blue}{asymptotic} minimax risk in estimating the
	location of a point source is lower bounded by
	\begin{equation} \label{eq:main-lower-bound}
		\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \frac{1}{32} \frac{\sigma^2 S}{nm\kappa^2w}.
	\end{equation}
\end{theorem}

\begin{figure}[t]
	\centering
	\includegraphics[width=2.8in]{overlap-middle-pics}
	\caption{Depiction of the continuous-space signals produced by two
		hypotheses, $\theta_0$ and $\theta_1$. Note that each signal is a
		shifted impulse response, and therefore has a support of size $w$.
		The set $\{\theta_0, \theta_1\}$ is a $2\delta$-packing of
		$\Theta$, so the signals are separated by a distance $2\delta$.}
	\label{fig:overlap-middle}
\end{figure}

\begin{figure}[t]
	\centering
	\includegraphics[width=2.8in]{delta-sampled-pics}
	\caption{Plot of the samples of the difference signal, $\Delta(k)$. The
		total size of the support of the difference signal is $w+2\delta$,
		hence at most $\floor*{\frac{(w + 2\delta)m}{S} + 1}$ samples of
		$\Delta(k)$ are non-zero.}
	\label{fig:delta-sampled}
\end{figure}

@@

sensors \textcolor{blue}{(plus 1, to account for edge-effects)} can lie in the region covered by

@@

example). \textcolor{blue}{In such a case, one part of the signal $x(s;\theta_0)$ will appear at

@@

repetition from the period $[S, 2S)$, at the right edge of the domain.} Also

@@

Combining equations~\eqref{eq:lb-after-pinsker} and~\eqref{eq:delta-bound},
\begin{equation} \label{eq:lb-after-delta-bound}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \frac{\delta^2}{2} \Bigg[1 - \sqrt{\frac{2n \kappa^2 \delta^2}{\sigma^2}\bigg(\!\frac{(w{+}2\delta) m}{S} + 1\!\bigg)} \Bigg].
\end{equation}
Asymptotically, we choose smaller values of $\delta$ as the number of sensors
grows large. Hence, \textcolor{blue}{neglecting edge effects} and terms of order $\delta^3$, we
tighten the bound by choosing $\delta$ as a function of the remaining variables
in order to have $\sqrt{\frac{2nm\kappa^2\delta^2 w}{\sigma^2 S}} =
\frac{1}{2}$. This is achieved for $\delta = \sqrt{\frac{\sigma^2
S}{8nm\kappa^2 w}}$, so that for $\Phi(\delta) = \delta^2$,
equation~\eqref{eq:lb-after-delta-bound} becomes
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \textcolor{blue}{\frac{1}{32}} \frac{\sigma^2 S}{nm\kappa^2 w}.
\end{equation}
This completes the proof.
\end{IEEEproof}

@@

The bound in Theorem~\ref{thm:main-lb} indicates that the error in source
localization goes to zero, as the number of sensors goes to infinity. This
observation is significantly different from the information-theoretic lower
bound given in~\cite{Grover2016Fundamental}, where even with an infinite number
of sensors, the lower bound is strictly greater than zero. This indicates that
the bound in section~\ref{sec:lecam-lb} is loose for large numbers of sensors.

The looseness lies in a poor sensor model: we assume that an arbitrarily large
number of sensors can be squeezed into a finite amount of space, and that these
sensors \emph{still} give us i.i.d.\ observations, with the same noise
variance. This is an impractical expectation. In reality, sensors have a
certain width, and the signal measured by them is the integral of the
continuous-space signal within that width (analogous to a sample-and-hold
circuit, which integrates a continuous-\emph{time} signal over a short period).
If the width of a sensor is decreased, then effectively, its signal-to-noise
ratio reduces (this is intuitively evident in the sample-and-hold analogy: to
compute the signal value, the integral must be divided by the time interval
over which the integration is performed; this division amplifies noise as the
time interval gets smaller).

@@

The bounding technique used in section~\ref{sec:lecam-lb} no longer easily
admits a solution which gives a bound for all values of $m$. We can compute the
bound numerically, however, using exact expressions for error probability in a
binary hypothesis test. For gaussian distributions of equal variance, the best
possible error probability (over all tests) is a Q-function in the distance
between their means:
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma(m)}\bigg)
\end{equation}
\textcolor{red}{Is this definitely correct? i.e.\ the best test is a linear
separator? I think this is coming from Neyman-Pearson. Should I cite?}
The minimax estimation risk is lower bounded by the risk in hypothesis testing,
hence
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) = \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma(m)}\bigg).
\end{equation}
This bound is valid for any pair of hypotheses $\theta_0$ and $\theta_1$,
separated by a distance of $2\delta$. As the $\delta$ increases, it becomes
easier to distinguish between the two hypotheses, so $P_e$ falls. When
numerically evaluating the bound, therefore, we have to \emph{manually}
trade-off between $\delta$ and $P_e$ to find the tightest possible bound. This
comes down to choosing $\theta_0$ and $\theta_1$ so that they are the hardest
hypotheses to distinguish between. To make the simulation easier, we can relax
the process of finding the least discernable hypothesis pair by fixing
$\theta_0$. Then, we need only optimize over $\delta$, making the process more
tractable. This comes at the cost of loosening the lower bound.

\begin{figure}[t]
	\centering
	\includegraphics[width=3.2in]{lb-vs-numsensors-int}
	\caption{A numerically generated plot of the lower bound on minimax risk of
	source localization, against the number of uniformly distributed sensors
used to estimate the location. The different curves are for different numbers
of ``trials'' (i.i.d.\ sensor observations). The bound saturates for large $m$,
indicating that for a fixed number of trials (or equivalently, for a fixed
SNR), there is reduced benefit in increasing the number of sensors beyond a
point. However, with more trials, the bound saturates at increasingly larger
numbers of sensors. Thus, there is benefit to increasing the number of sensors,
provided SNR can be increased by leveraging larger numbers of trials.}
	\label{fig:numerical}
\end{figure}

Plotting this numerically computed bound for different numbers of trials (see
Fig.~\ref{fig:numerical}) reveals some interesting trends. First, the bound
saturates as $m$ grows large, matching the behaviour presented
in~\cite{Grover2016Fundamental}. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. With fewer sensors, it is harder to
estimate the location of the point source. Third and most importantly, the
point at which the bound begins to saturate appears to increase when we use
more trials. Recall that the number of ``trials'' is the number of i.i.d.\
sensor observations we receive. Intuitively, estimation is eased with more
trials, as we could perform more averaging, and increase SNR. \emph{With higher
SNR, there is greater benefit in using a larger number of sensors.} This
matches the intuition given in~\cite{Grover2016Information}, where it is
propounded that using more sensors allows us to sample a greater number of
high-frequency components of the source signal, and hence get a better estimate
of its location.

>>>>

To be considered for the 2017 IEEE Jack Keil Wolf ISIT Student Paper Award.

@@

The sensors sample the continuous-space shifted impulse response, with some
additive noise. We denote the noiseless sampled version of $x(s; \theta)$ by
the $m$-length vector $\v x(\theta)$:
\begin{equation} \label{eq:sampled-signal}
	\v x(\theta) = \bigg[x(0; \theta), \ldots, x\Big(\frac{kS}{m}; \theta\Big), \ldots, x\Big(\frac{(m-1)S}{m}; \theta\Big)\bigg]^T
\end{equation}
where $m$ is the number of sensors and $k \in \{0, 1, \ldots, m-1\}$. The
additive noise is given by $\v \epsilon$ (to be described shortly), and the
noisy samples are represented by
\begin{equation} \label{eq:sensor-obs}
	\v y = \v x(\theta) + \v \epsilon.
\end{equation}
In the source localization problem, we receive $n$ trials\footnote{The word
``trials'' comes from neuroscience literature, where several trials of an
experiment are run with a single participant. The participant is given the same
stimulus in each trial, evoking the same response in the brain (usually with
some trial-to-trial variability). For our purposes, we assume that trials are
effectively i.i.d. realizations of sensor noise, with no trial-to-trial
variability.} (i.e., $n$ realizations) of $\v y$ according to
equation~\eqref{eq:sensor-obs}, for the same point source $\theta$, using which
we need to estimate the location of the source.

@@

The noise $\v \epsilon$, introduced in equation~\eqref{eq:sensor-obs}, is the
only source of uncertainty in the problem. For our purposes, we assume that $\v
\epsilon$ is zero-mean and Gaussian, $\v \epsilon \sim \mathcal{N}(\v 0, \m
\Sigma)$. We restrict our analysis to a ``sensor noise'' setting, wherein
$\m\Sigma = \sigma^2 \m I$, i.e., noise is i.i.d.\ across sensors.

@@

\textcolor{blue}{We start by lower bounding the maximum risk of the estimation problem with the
average risk of a multiple hypothesis testing problem.} For this, we first need
to define a $2\delta$-packing:%
\begin{definition}
	A set $\Theta_{\mathcal{V}} = \{ \theta_v : v \in \mathcal{V} \}$ for some
	finite index set $\mathcal{V} \subset \mathbb N$ is said to be a
	$2\delta$-packing in the $\rho$-metric if
	\begin{equation}
		\rho(\theta_i, \theta_j) \geq 2\delta \qquad \forall \;\; \theta_i, \theta_j \in \Theta_{\mathcal{V}}.
	\end{equation}
\end{definition}
\begin{theorem} \label{thm:est-to-testing}%
	If we can find a $2\delta$-packing $\Theta_{\mathcal{V}}$ of $\Theta$, then
	we can lower bound the minimax estimation risk by the average testing risk:
	\begin{equation}
		\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) \inf_\psi \mathbb P (\psi(Y^n) \neq V)
	\end{equation}
	where $V$ is the unknown true hypothesis which takes values uniformly from
	$\mathcal{V}$, and $\psi$ is our estimate of the hypothesis.
\end{theorem}
For a proof of this theorem, we refer the reader to Proposition~13.3
in~\cite{Duchi2015Information}.  Intuitively, Theorem~\ref{thm:est-to-testing}
says that the error in estimating $\theta_i$ is likely to be large if it is
difficult to distinguish $\theta_i$ from $\theta_j$, i.e.\ if the probability
of error is high.

@@

For the source localization problem, we measure \emph{risk} in terms of the
squared-distance between the true location and the estimated location. Hence,
our loss function is $\Phi(\rho(\theta, \hat\theta)) = \abs{\theta -
\hat\theta}^2$, where $\hat\theta$ is some estimate of the location based on
$n$ trials of the noisy sensor observations $\v y$.

\begin{figure}[t]
	\centering
	\includegraphics[width=2.6in]{overlap-middle-pics}
	\caption{Depiction of the continuous-space signals produced by two
		hypotheses, $\theta_0$ and $\theta_1$. Note that each signal is a
		shifted impulse response, and therefore has a support of size $w$.
		The set $\{\theta_0, \theta_1\}$ is a $2\delta$-packing of
		$\Theta$, so the signals are separated by a distance $2\delta$.}
	\label{fig:overlap-middle}
\end{figure}
\begin{figure}[t]
	\centering
	\includegraphics[width=2.6in]{delta-sampled-pics}
	\caption{Plot of the samples of the difference signal, $\Delta(k)$. The
		total size of the support of the difference signal is $w+2\delta$,
		hence at most $\floor*{\frac{(w + 2\delta)m}{S} + 1}$ samples of
		$\Delta(k)$ are non-zero.}
	\label{fig:delta-sampled}
\end{figure}

We now state the main result of this paper.
\begin{theorem} \label{thm:main-lb}
	For a source localization problem as defined in
	section~\ref{sec:source-localization}, the minimax risk in estimating the
	location of a point source is lower bounded by
		\mathfrak{M}_n(\mathcal{P}, \Phi\circ\rho) \geq \sup_{0 < \delta < S/4} \frac{\delta^2}{2} \Bigg[1 - \sqrt{\frac{2n \kappa^2 \delta^2}{\sigma^2}\bigg(\!\frac{(w{+}2\delta) m}{S} + 1\!\bigg)} \Bigg]
	For sufficiently large $m$, the bound can be approximated by
	\begin{equation} \label{eq:main-lower-bound}
		\mathfrak{M}_n(\mathcal{P}, \Phiorho) \gtrsim \frac{1}{32} \frac{\sigma^2 S}{nm\kappa^2w}.
\end{theorem}%

@@

sensors (plus 1, to account for edge-effects) can lie in the region covered by

@@

example). In such a case, one part of the signal $x(s;\theta_0)$ will appear at

@@

repetition from the period $[S, 2S)$, at the right edge of the domain. Also

@@

Combining equations~\eqref{eq:lb-after-pinsker} and~\eqref{eq:delta-bound},
\begin{equation} \label{eq:lb-after-delta-bound}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \frac{\delta^2}{2} \Bigg[1 - \sqrt{\frac{2n \kappa^2 \delta^2}{\sigma^2}\bigg(\!\frac{(w{+}2\delta) m}{S} + 1\!\bigg)} \Bigg].
\end{equation}
This bound holds for any feasible $2\delta$-packing. We can derive the tightest
possible bound by maximizing the above quantity over all feasible $\delta$.
Since the maximum attainable separation between two hypotheses on the periodic
domain is $S/2$, we maximize $\delta$ over $(0, S/4)$. This yields the theorem.

To derive the approximation, we observe that for sufficiently large numbers of
sensors, the optimal value of $\delta$ starts becoming small, as hypotheses
must chosen to be closer in order to make the error probability large. Hence,
neglecting edge effects and terms of order $\delta^3$, we can give an
approximate bound for large $m$ using a heuristic: tighten the bound by
choosing $\sqrt{\frac{2n\kappa^2\delta^2wm}{\sigma^2 S}} = \frac{1}{2}$. This
is achieved for $\delta = \sqrt{\frac{\sigma^2 S}{8nm\kappa^2 w}}$, so that for
$\Phi(\delta) = \delta^2$, equation~\eqref{eq:lb-after-delta-bound} becomes
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \frac{1}{32} \frac{\sigma^2 S}{nm\kappa^2 w}.
\end{equation}
This completes the proof.
\end{IEEEproof}

@@

The bound in Theorem~\ref{thm:main-lb} may mislead one to think that the error
in source localization goes to zero, as the number of sensors goes to infinity.
This observation is qualitatively different from the information-theoretic
lower bound given in~\cite{Grover2016Fundamental}, where even with an infinite
number of sensors, the lower bound is strictly greater than zero. This
indicates that the bound in section~\ref{sec:lecam-lb} is failing to capture
some effect which leads to saturation of error, as the number of sensors
increases.

Indeed, the bound in Theorem~\ref{thm:main-lb} assumes that an arbitrarily
large number of sensors can be squeezed into a finite amount of space, and that
these sensors \emph{still} give us i.i.d.\ observations, with the same noise
variance. This is aphysical, since in reality, sensors have a certain width,
and the signal measured by them is the integral of the continuous-space signal
within that width (analogous to a sample-and-hold circuit, which integrates a
continuous-\emph{time} signal over a short period).  If the width of a sensor
is decreased, then effectively, its signal-to-noise ratio reduces. This is
intuitively evident in the sample-and-hold analogy: to compute the signal
value, the integral must be divided by the time interval over which the
integration is performed; this division amplifies noise as the time interval
gets smaller.

@@

The bounding technique used in section~\ref{sec:lecam-lb} no longer easily
admits a solution which gives a bound for all values of $m$. We can compute the
bound numerically, however, using exact expressions for error probability in a
binary hypothesis test. For gaussian distributions of equal variance and a
uniform prior on the hypotheses, the best possible error probability (over all
tests) is a Q-function~\cite{Qfunction} in the distance between their
means~\cite{NPLemmaOrSimilar},
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma(m)}\bigg).
\end{equation}
The minimax estimation risk is lower bounded by the risk in hypothesis testing,
hence
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma(m)}\bigg).
\end{equation}
This bound is valid for any pair of hypotheses $\theta_0$ and $\theta_1$,
separated by a distance of $2\delta$. As the $\delta$ increases, it becomes
easier to distinguish between the two hypotheses, so $P_e$ falls. When
numerically evaluating the bound, therefore, we have to \emph{manually}
trade-off between $\delta$ and $P_e$ to find the tightest possible bound. This
comes down to choosing $\theta_0$ and $\theta_1$ so that they are the hardest
hypotheses to distinguish between. To make the simulation easier, we can relax
the process of finding the least discernable hypothesis pair by fixing
$\theta_0$. Then, we need only optimize over $\delta$, making the process more
tractable. This comes at the cost of loosening the lower bound.

\begin{figure}[t]
	\centering
	\includegraphics[width=3.2in]{lb-vs-numsensors-int}
	\caption{A numerically generated plot of the lower bound on minimax risk of
		source localization, against the number of uniformly distributed
		sensors, $m$, used to estimate the location. The different curves are
		for different numbers of ``trials'' $n$ (i.i.d.\ sensor observations).
		The bound saturates for large $m$, indicating that for a fixed number
		of trials (or equivalently, for a fixed SNR), there is reduced benefit
		in increasing the number of sensors beyond a point. However, with more
		trials, the bound saturates at increasingly larger numbers of sensors.
	Thus, there is benefit to increasing the number of sensors, provided SNR
can be increased by leveraging larger numbers of trials.}
	\label{fig:numerical}
\end{figure}

Plotting this numerically computed bound for different numbers of trials (see
Fig.~\ref{fig:numerical}) reveals some interesting trends. First, the bound
saturates as $m$ grows large, matching the behaviour presented
in~\cite{Grover2016Fundamental}. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. With fewer sensors, it is harder to
estimate the location of the point source. Third and most importantly, the
point at which the bound begins to saturate increases when we use
more trials. Recall that the number of ``trials'' is the number of i.i.d.\
sensor observations we receive. Intuitively, estimation is eased with more
trials, as we could perform more averaging, and increase SNR. \emph{With higher
SNR, there is greater benefit in using a larger number of sensors.} This
matches the intuition given in~\cite{Grover2016Information}, where it is
propounded that using more sensors allows us to sample a greater number of
high-frequency components of the source signal, and hence get a better estimate
of its location.
